Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap

In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and
\(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the
subcategory consisting only of the associators, unitors, and symmetry, and only
the identity object considered affine and relevant. On the other extreme, every
\textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural
Cartesian category with all morphisms both relevant and affine; in fact, this is
the case \textit{if and only if} a category is Cartesian, hence the term
"substructural Cartesian category."

We can also deduce the following:
\begin{proposition}
  If \(\mc{C}\) is a substructural monoidal category then the Kleisli category
  \(\mc{C}_{\mb{T}}\) is a substructural premonoidal category with:
  \begin{itemize}
    \item \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\), \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\)
    \item \(\mc{C}_{\mb{T}}^q(A, B) = \upg{\mc{C}^q(A, B)}{}\), where
    \(\upg{f}{} = f;\eta\)
    \item Drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\)
  \end{itemize} 
  In particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a
  substructural effectful category.
\end{proposition}

\TODO{work into text?}

@@

We begin with some definitions:
\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a
  monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to
  I\) mapping (infinite) streams of \(M\) to elements of \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the
  stream \(\ell\).
\end{definition}
Some basic examples include:
\begin{enumerate}
  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
  forms a stream action on the one element set \(\mathbf{1}\).
  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
  stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
  \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
  compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
  \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
  \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
  \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
  \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
  \cdot_i l_i)_i\).
  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
  \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite)
  sequence \(s\).
  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
  stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
  on \(\nats \cup \{\infty\}\) by addition.
\end{enumerate}
We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
\(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
\omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) =
\{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains
the unit of \(M\).

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the
  \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
  \times M + I)\), with \(\eta_A = \lambda a.\iota_0 (a, 1)\) and 
  \[
    \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
  \]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an
  Elgot monad, with iteration operator \(f^\dagger(a) = f^\infty(a) \cup
  \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
  \in f_n(a)\}\), where
  \[
    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
  \]
  \[
    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
  \]
  \TODO{clean this up}
\end{definition}

@@

  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)

>>>>

In general, \textit{every} premonoidal category \(\mc{C}\) can be viewed a
substructural premonoidal category, with \(\mc{C}^\varnothing = \mc{C}\) and
\(\mc{C}^{\{\ms{a}, \ms{r}\}}, \mc{C}^{\{\ms{a}\}}, \mc{C}^{\{\ms{r}\}}\) the
subcategory consisting only of the associators, unitors, and symmetry, and only
the identity object considered affine and relevant. Similarly, every
\textit{Cartesian} category \(\mc{C}\) can be viewed as a substructural
Cartesian category with all morphisms both relevant and affine; in fact, this is
the case \textit{if and only if} a category is Cartesian, hence the term
"substructural Cartesian category." We also note the following:
\begin{proposition}
  If \(\mc{C}\) is a substructural monoidal category then we can equip the
  Kleisli category \(\mc{C}_{\mb{T}}\) with the structure of a substructural
  effectful category by taking \(\ms{Aff}(\mc{C}_{\mb{T}}) = \ms{Aff}(\mc{C})\),
  \(\ms{Rel}(\mc{C}_{\mb{T}}) = \ms{Rel}(\mc{C})\), and \(\mc{C}_{\mb{T}}^q(A,
  B) = \upg{\mc{C}^q(A, B)}{}\), where \(\upg{f}{} = f;\eta\), and defining
  drops \(\upg{\ms{drop}(A)}{}\) and splits \(\upg{\ms{split}(A)}{}\). In
  particular, \(\upg{\cdot}{}: \mc{C} \to \mc{C}_{\mb{T}}\) is a substructural
  effectful category.
\end{proposition}

@@

We begin with some definitions:
\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a
  monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to
  I\) mapping (infinite) streams of \(M\) to elements of \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the
  stream \(\ell\).
\end{definition}
Some basic examples include:
\begin{enumerate}
  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
  forms a stream action on the one element set \(\mathbf{1}\).
  % \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
  % \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
  % \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
  % \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
  % \cdot_i l_i)_i\).
  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream
  action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\),
  the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell
  \cdot s\) prepends the \(\ell\) to \(s\).
  % \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
  % stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
  % on \(\nats \cup \{\infty\}\) by addition.
  \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
  stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
  \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
  compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
\end{enumerate}
We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
\(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
\omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) =
\{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains
the unit of \(M\).

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the
  \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
  \times M + I)\), with \(\eta_A = \lambda a.\iota_0 (a, 1)\) and 
  \[
    \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
  \]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). The iteration operator \(f^\dagger(a) = f^\infty(a) \cup
  \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
  \in f_n(a)\}\) yields an Elgot structure, where
  \[
    f_0 = f, \quad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n], \quad
    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
  \]
\end{definition}

@@

  \item Relevant and affine morphisms to be the lifts of relevant/affine
  morphisms in \(\ms{Print}\)
